\documentclass[a4paper]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{blindtext}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{booktabs}
\usepackage{float}
\usepackage{bbm}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows}
% \usepackage{stix}
\graphicspath{ {./images/} }
\usepackage[bottom=0.5cm, right=1.5cm, left=1.5cm, top=1.5cm]{geometry}

\newtheorem{theorem}{Theorem}
\newtheorem{exercise}{Exercise}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}[section]

\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}
\newcommand{\p}{\ensuremath{\mathbb{P}}}
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\1}{\ensuremath{\mathbbm{1}}}
\newcommand{\B}{\ensuremath{\mathbbm{B}}}

\title{Lecture 12}
\author{Gidon Rosalki}
\date{2025-06-22}


\begin{document}
\maketitle
\begingroup
\centering
{\color{red}\bfseries\LARGE This content will not appear in the exam this year! \par}
\endgroup

\vspace{1cm}  % Add space before the rest of the content
We are going to discuss the classes NL, coNL, applying Savitch's theorem on NL (and sublinear classes), and Immerman's
theorem NL = coNL, and use and proof.
\section{Sublinear space classes}\label{sec:Sublinear space classes} % (fold)
\begin{theorem}[Savitch]
    For every function $f \left(n\right)$ computable in space $O \left(f \left(n\right)\right)$ it holds that
    \[
        NSpace \left(f \left(n\right)\right) \subseteq Space \left(f^2 \left(n\right)\right)
    \]
    We showed this for every $n \leq f \left(n\right)$, and will now show this for $\log \left(n\right) \leq f
    \left(n\right) \leq n$
    \begin{proof}[Proof ]
        Let there be a language $A \in \text{NSPACE} \left(f \left(n\right)\right)$, and let there be a NTM $M$, that
        recognises $A$, in space $O \left(f \left(n\right)\right)$. For a word $w$, we will look at the configurations
        graph $G_{M, w}$, which has a node for each configuration, an edge for each possible transition of $\delta$ of
        $M$, and two special configurations $c_0$, the starting configuration, and $c_{\text{acc}} $, the accepting
        configuration. $w \in A$ \textbf{if and only if} There exists a run that accepts $M \left(w\right)$ \textbf{if
        and only if} $\left(G_{M, w}, c_0, c_{\text{acc}} \right) \in \text{PATH} $. That is to say, there is a directed
        path from $c_0$, to $c_{\text{acc}} $ in $G_{M, w}$. \\
        We have built a DTM $M'$ that \textbf{decides} if $\left(G_{M, w}, c_0, c_{\text{acc}} \right) \in \text{PATH}
        $, while only using a relatively small amount of space $O \left(f^2 \left(n\right)\right)$. The full proof is
        located in a previous lecture. \\
        $M'$ passes over every middle node in the graph (all the nodes). If there exists a path for this 3-tuple of
        nodes, then it returns true. If not, it returns false, and moves on to the next tuple. Correctness is trivial.
        The space compleity is the sum of the complexity of 3 nodes, times the depth of the recursion. This is the
        log of number of nodes in the graph, so overall we get $O \left(\log^2 \left(\text{num nodes in graph}
        \right)\right)$. The number of the nodes in the graph is \[
                \left|Q\right| \cdot n \cdot O \left(f \left(n\right)\right) \cdot \left|\Gamma\right|^{O \left(f
                \left(n\right)\right)}
        \]
        So, since $\log \left(\# \text{conf} \right) = O \left(f \left(n\right)\right)$, in total the machine $M'$ uses
        $O \left(f^2 \left(n\right)\right)$ space.

    \end{proof}
\end{theorem}

What is the number of configurations for a machine that is NL? (Where $f \left(n\right) = O \left(\log
\left(n\right)\right)$ \[
    \# \text{conf} = \left|Q\right| \cdot \left(n + 1\right) \cdot O \left(\log \left(n\right)\right) \cdot
    \left|\Gamma\right|^{O \left(\log \left(n\right)\right)}
\]
Which is comprised of the internal states, the head location on the input tape, the location of the head on the work
tape, and the content of the work tape.

Note: We will note that in the above count, we did not count the possibilities of the input tape (which are
$\left|\Sigma\right|^n$), since $w$ is given, which is to say does not vary between different nodes of the graph $G_{M,
w}$, we did not count it.

So in total we get \[
    \log \left(\# \text{conf} \right) = O \left(\log \left(n\right)\right)
\]
% section Sublinear space classes (end)


\section{Immerman's theorem}\label{sec:Immerman's theorem} % (fold)
\begin{theorem}[]
    \[
        \text{NL}  = \text{coNL}
    \]
    Which is to say \[
        \text{NSpace}  \left(\log \left(n\right)\right) = \text{coNSpace}  \left(\log \left(n\right)\right)
    \]
    Or \[
        A \in \text{NL} \Leftrightarrow \overline{A} \in \text{NL}
    \]

\end{theorem}
We have already seen that NPSpace = PSpace. Reminder: The fact that PSpace $\subseteq$ NPSpace is trivial. The
other direction, NPSpace $\subseteq$ PSpace can be shown through Savitch's theorem. Let there be $A \in
\text{NPSpace} = \displaystyle\bigcup_{k = 1}^{\infty}\text{NSpace} \left(n^k\right)$. That is to say, there
exists $k$ such that $A \in \text{NSpace} \left(n^k\right)$. From Savitch, it holds that $A \in \text{Space}
\left(\left(n^k\right)^2\right) = \text{Space} \left(n^{2k}\right)$, and so $A \in \displaystyle\bigcup_{k =
1}^{\infty}\text{Space} \left(n^k\right) = \text{PSpace} $.

Let us move on to our actual theorem, NPSPACE = coNSPACE, which is to say $A \in \text{NPSPACE} \Leftrightarrow
\overline{A} \in \text{NPSPACE} $. \\
Proof: Let there be $A \in \text{NPSPACE} $. We will show that $\overline{A} \in \text{NPSPACE} $. $A \in
\text{NPSPACE}$, however NPSPACE = PSPACE, so $A \in \text{PSPACE} $. This is to say that there exists a DTM that
decides $A$, in polynomial space, and so it is also true that $\overline{A} \in \text{PSPACE} $, since the same DTM that
decides $A$ can decide $\overline{A}$ by swapping its accepting and rejecting states, with the same space usage, and so
it also holds that $\overline{A} \in \text{NPSPACE} $. \\

Can we use the same method to show that NL = coNL?
No. Using Savitch in order to transition from non-deterministic to deterministic increases the space usage to its
square. For some space that is polynomial, that matters not, but for NL, this gives us \[
    B \in NL \implies B \in \text{SPACE }\left(\log^2 \left(n\right)\right)
\]
It s not clear if from here we can return to coNL, or NL, it is possible to achieve a DTM that decides $\overline{B}$,
but it will require $O \left(\log^2 \left(n\right)\right)$.

Let us try a different tack. Given $A \in $ coNL, and so $\overline{A} \in $ NL, let there be a machine NL $M$ for
$\overline{A}$. Can we create an NL machine for $A$ by swapping the states $q_{\text{acc}}, q_{\text{rej}} $? We cannot.
This is becuse the machine NL for $\overline{A}$ ensures that if $w \in \overline{A}$, there is an accepting run, and if
$w \notin A$, every run is rejected. Swapping the states $q_{\text{acc}}, q_{\text{rej}} $ will give a machine that if
$w \in \overline{A}$, there is a run that rejects, and if $w \notin \overline{A}$ every run is accepted, and so if $w
\notin A$, there is a run that accepts, and if $w \in A$, every run will accept. As we can see, this does not define the
machine that we want.

Concrete example: Let us use \[
    \text{PATH} = \left\{\left(G, s, t\right) : \text{There is a directed path from $s$ to $t$ in } G\right\}
\]
We saw that $\text{PATH} \in \text{NL-Complete} $. Furthermore, $\text{PATH} \in \text{NL} $. We saw an NL machine for
PATH, that beings from $s$, and non deterministically chooses a path, and returns $q_{\text{acc}} $ \textbf{if and only
if} the path finishes at $t$. Swapping the states $q_{\text{acc}}, q_{\text{rej}} $ creates a machine that does not
ensure correctness for $\overline{\text{PATH} }$. For $\left(G, s, t\right)$, if the machines returns $q_{\text{acc}} $,
it is because there is not a path from $s$ to $t$ in $G$, which is to say that $\left(G, s, t\right) \notin
\overline{\text{PATH} }$. However, given that the machine found a single path, there is a path from $s$ to $t$ in $G$,
which is to say that the machine was meant to return $q_{\text{rej}} $.

\begin{theorem}[]
    \[
        NL = coNL \Leftrightarrow \overline{\text{PATH} } \in NL
    \]
    \begin{proof}[Proof ]
        $NL = coNL \implies \overline{\text{PATH} } \in NL$. If $NL = coNL$, then since PATH $\in $ NL, we can get that
        $\overline{\text{PATH} } \in NL$ \\

        $\overline{\text{PATH} } \in NL \implies NL = coNL  $. Reminder, PATH $\in $ NL-Hard, let $A \in coNL$, then
        $\overline{A} \in \text{NL} $. So, \[
            \overline{A} \leq_L \text{PATH}
        \]
        and therefore \[
            A \leq_L \overline{\text{PATH} }
        \]
        through the same reduction. Since it is given that \[
            \overline{\text{PATH} } \in \text{NL}
        \]
        there \[
            A \in \text{NL}
        \]
        by the reduction theorem of language in $NL$. In conclusion \[
            \text{coNL} \subseteq \text{NL}
        \]
        However, we wanted equivalence, not containment. Let there be $B \in \text{NL} $, therefore \begin{gather*}
            B \in \text{NL}  \\
            \implies \overline{B} \in coNL \\
            \text{According to the containment} \implies \overline{B} \in NL \\
            \implies NL \subseteq coNL \\
            \implies NL = coNL
        \end{gather*}
        In conclusion, $\overline{PATH} \in NL$ is a necessary requirement, and sufficient to show that \[
            NL = coNL
        \]
        Therefore, in order to prove Immerman's theorem, we want to show an NL machine for
        \[
            \overline{\text{PATH} } = \left\{\left(G, s, t\right) : \text{ In the directed graph $G$, there is no path from $s$
            to $t$} \right\}
        \]
    \end{proof}
\end{theorem}

\subsection{Examples for use of Immerman}\label{sub:Example for use of Immerman} % (fold)
\begin{example}[]
    \[
        A = \left\{\left(G, u, v\right) : w, v \text{ are not in the same strongly connected component in the directed
        graph } G\right\}
    \]
    Reminder: 2 nodes $u, v$ are in the same strongly connected component \textbf{if and only if} there is a directed
    path from $u$ to $v$, and also from $v$ to $u$ in $G$. \\
    Reminder: If $C, D \in NL$, then so too are $C \cap D$ and $C \cup D$.

    \begin{proof}[Solution]
        We need to show that $A \in \text{NL} $: \\
        $\overline{\text{PATH} } \in \text{NL} $ from Immerman, and
        \begin{align*}
            A \in \left(G, y, v\right) &\Leftrightarrow \text{There is no path from $u$ to $v$ or there is no path from
            $v$ to $u$}  \\
                                       &\Leftrightarrow \left(G, u, v\right) \in \overline{\text{PATH} } \lor \left(G,
                                       v, u\right) \in \overline{\text{PATH} }
        \end{align*}
        That is to say, the union of the two language in NL, which is to say that $A$ is in NL.
    \end{proof}
\end{example}

\begin{example}[]
    \[
        \text{2-Con} = \left\{G : G \text{ is a directed graph, and has at least 2 strongly connected components} \right\}
    \]

    \begin{proof}[Solution]
        We need to show that 2-con $\in $ NL: \\
        We will create an NL-machine for 2-Con: $M$ will run as follows: For every pair of nodes $u, v \in V
        \left(G\right)$, it will run the machine of $\overline{\text{PATH} }$ on $\left(G, u, w\right)$. If the machine
        returns $q_{\text{acc}} $, then $M$ will also return such, and stop. Otherwise, $M$ will move onto the next pair
        $u, v$. If $M$ has tried all the pairs $u, v$, then it will return $q_{\text{rej}} $. \\

        Correctness: If $G \in $ 2-Con, then there exists a pair of nodes $u, v$ that for not in the same strongly
        connected component. W.l.o.g. there is no path from $u$ to $v$, and so when running the machine
        $\overline{\text{PATH} }$ on $\left(G, u, v\right)$, there is a run that returns $q_{\text{acc}} $, therefore
        there is a run of $M$ that returns $q_{\text{acc}} $. \\
        If $G \notin $ 2-con, which is to say all of $G$ is a single strongly connected component, then for every pair
        of nodes $u, v$ every run of the machine $\overline{\text{PATH} }$ on $\left(G, u, v\right)$ returns
        $q_{\text{rej}} $, and therefore $M$ always returns $q_{\text{rej}} $.

        Space complexity: A variable for $u$, and for $v$: $O \left(\log \left(n\right)\right)$, the space for running
        $\overline{\text{PATH} }$: $O \left(\log \left(n\right)\right)$, in total $O \left(\log \left(n\right)\right)$.
    \end{proof}
\end{example}
% subsection Example for use of Immerman (end)

\subsection{Proof of Immerman}\label{sub:Proof of Immerman} % (fold)
We need to show that $\overline{\text{PATH} } \in $ NL. \\
\subsubsection{General idea}\label{sec:General idea} % (fold)
We will look at NL machines that carry out computations, and ensure that during the run: \begin{enumerate}
    \item The function that we compute $f \left(n\right)$ is written in the output, and the internal state is
        $q_{\text{acc}} $
    \item Or we are not sure what is on the output, and the internal state is $q_{\text{rej}} $
    \item For every input there is at least 1 run which finishes in $q_{\text{acc}} $
\end{enumerate}
How does this help? Consider the following example. \\
Let us assume that there is a machine NL, as described just above (three requirements) that for the input $\left(G,
s\right)$, computes $c$, the number of nodes that can be reached from $s$ in $G$. We will call this machine $M'$. With
$M'$, we may build an NL machine for $\overline{\text{PATH} }$. \[
    \overline{\text{PATH} } = \left\{\left(G, s, t\right) : \text{ In the directed graph $G$, there is no path from $s$
    to $t$} \right\}
\]
We will run $M'$, once on $\left(G, s\right)$, and compute $c$, and a second time on $\left(G \setminus
\left\{t\right\}, s\right)$, and compute $c'$. That is to say, the second time we run $M'$, and compute how many nodes
can be reached from $s$, when we remove $t$ from $G$. If $c = c'$, then $\left(G, s, t\right) \in \overline{\text{PATH}
} $, and if $c \ne c'$, the $\left(G, s, t\right) \in \text{PATH} $. $M$ will therefore return $q_{\text{acc}} $
\textbf{if and only if} $c = c'$.

We will use $c_k$ to be the number of nodes that one can reach from $s$ in $G$ by $k$ steps or less. So \begin{itemize}
    \item $c_0 = 1$, in a path of length $\leq 0$, we can only reach $s$
    \item $c_1 = $ the number of neighbours of $s$ + 1. In a path of length $\leq 1$, we can reach $s$, and all its
        immediate neighbours
\end{itemize}
We will show a method to compute $c_{k + 1}$ by using $\left(G, s, c_k\right)$. The machine $M''$ will run as follows:
Beginning from $c_0 = 1$, and compute in a loop until $c_n$ where $n$ is the number of edges in $G$. Similarly, we will
run $M''$ in order to compute $c_0'$, until $c_n'$ for $G'$, where the graph is like $G$, but without the node $t$. The
machine $M$ will compute $c_n$ and $c_n'$, and answer $q_{\text{acc}} $ \textbf{if and only if} $c_n = c_n'$. \\
NOTE: The machine $M''$ may transfer to $q_{\text{rej}} $ at every stage of a run,
but if $\left(G, s, t\right) \in \overline{\text{PATH} }$, then there is a run such that $M''$ does not return
$q_{\text{rej}} $ at all. That is to say, correctly computes the values $c_k$. We will focus on the task of computing
$c_{k + 1}$ from $c_k$. \\
For every node $v_1, \dots, v_n$ we want to know with \textbf{certainty} if we can reach it via a path of length less
than equal to $k + 1$ . If so, we will increase the counter $c_{k + 1}$ by 1, and if not, then we will not. At the end
of the process, we will get a justified value for $c_{k + 1}$.

How will will know with certainty if it is possible to reach $v_1$, within $k + 1$ steps? It is possible \textbf{if and
only if} there is a node $v_i$ which can be reached in $k$ or less steps, and from there there is an edge to $v_1$.
Therefore, for every node $v_i \in V \left(G\right)$, it is the candidate to be the $v_i$ we want to reach from $s$
within $k$ steps, and we will compute the path from $s$ to $v_i$ in length $\leq k$. If we succeed to reach $v_i$, then
we will increase the counter $D_k$ by 1, and check if there is an edge $\left(v_1, v_i\right)$. After we pass over all
the edges for $v_i$, we will check if $D_k = c_k$. \begin{itemize}
    \item If so - We will note that we have passed over all the nodes of distance $\leq k$
    \item If not - we have missed at least 1
\end{itemize}
Therefore, if so, and we have not managed to arrive to $v_1$, then we can be certain that we cannot reach it. However,
if not, which is to say $D_k \ne c_k$, we will stop and return $q_{\text{rej}} $. Similarly to $v_1$, we will do this
for every node $v_2, \dots, v_n$, of course making sure to zero $D_k$ every time.

Space usage: For $D_k, c_k, c_{k + 1}$ we need $O \left(\log \left(n\right)\right)$, and for the guessed path from $s$,
we need $O \left(\log \left(n\right)\right)$, because we guess a path step by step, and count how many steps there were.
For $c, c'$, we need $O \left(\log \left(n\right)\right)$. In total, $O \left(\log \left(n\right)\right)$. \\
Correctness: For every $k$, if we did not return $q_{\text{rej}} $, we get that $c_{k + 1}$ was computed correctly from
$c_k$. Therefore, both $c$, and $c'$ are correctly computed (in a run where we do not return $q_{\text{rej}} $), which
is to say $M$ correctly solves $\overline{PATH}$
% subsubsection General idea (end)
% subsection Proof of Immerman (end)
% section Immerman's theorem (end)

\end{document}
