\documentclass[a4paper]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{blindtext}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{booktabs}
\usepackage{float}
\usepackage{bbm}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows}
% \usepackage{stix}
\graphicspath{ {./images/} }
\usepackage[bottom=0.5cm, right=1.5cm, left=1.5cm, top=1.5cm]{geometry}

\newtheorem{theorem}{Theorem}
\newtheorem{exercise}{Exercise}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}[section]

\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}
\newcommand{\p}{\ensuremath{\mathbb{P}}}
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\1}{\ensuremath{\mathbbm{1}}}
\newcommand{\B}{\ensuremath{\mathbbm{B}}}

\title{Lecture 11}
\author{Gidon Rosalki}
\date{2025-06-15}


\begin{document}
\maketitle
\section{Sub linear space complexity}\label{sec:Sub linear space complexity} % (fold)
In order to discuss sub linear space classes, we defined TMs with 3 tapes: \begin{enumerate}
    \item Input: read only
    \item Work tape: read / write, we use only this tape to measure space
    \item Output: Write only, at every step may write a letter, or not. If we write, then we move the head one step to
        the right, and never return left.
\end{enumerate}

We defined the classes \begin{align*}
    L &= Space \left(\log \left(n\right)\right) \\
    NL &= NSpace \left(\log \left(n\right)\right)
\end{align*}

\begin{theorem}[Savitch]
    \[
        NSpace \left(f \left(n\right)\right) \subseteq Space \left(f^2 \left(n\right)\right)
    \]
    For every $f \left(n\right)$ that is computable in $O \left( f \left(n\right)\right)$.
    \begin{proof}[Proof]
        We showed this for functions $f \left(n\right) \geq n$, and the theorem is true also for functions $f
        \left(n\right) < n$
    \end{proof}

\end{theorem}

Let there be $f \left(x\right)$, a computable function in space $O \left(\log \left(n\right)\right)$. That is to say,
there is a TM $M_f$, with 3 tapes (as defined above), such that for every input $x$ on the input tape. $M_f$ finished in
state $q_{\text{acc}} $, with $f \left(x\right)$ on the output tape, while using space $O
\left(\log\left(n\right)\right)$ on the work tape, where $n = \left|x\right|$. Let $g \left(n\right)$ be a computable
function, in
space $O \left(\log \left(n\right)\right)$, by the TM $M_g$.

\begin{theorem}[]
    $f \left(g \left(x\right)\right)$ is computable in space $O \left(\log \left(n\right)\right)$.

    \begin{proof}[Proof ]
        We showed that if $f$ and $g$ are computable, then $f \circ g$ is computable, and we showed that if $f$ and $g$
        are computable in polynomial time, then so too $f \circ g$ are computable in polynomial time, by creating a new
        TM $M$, which for an input $x$, runs $M_g$ on $x$, so $w = M_g \left(x\right)$, and saves $w$ on the tape, and
        then runs $M_f$ on $w$, so thus we receive $f \left(g \left(x\right)\right)$, which is clearly computable, since
        each individual function is computable, and clearly occurred in polynomial time, since both individual
        computations occurred in computational time.

        How can we create 2 machines, both with 3 tapes, by using a machine with 3 tapes, that computes $f \left(g
        \left(x\right)\right)$? \\
        Solution 1: Correct, but wasteful of space: We will change $M_g$ such that instead of printing to the output, it
        will print to the work tape, and we will change $M_f$, such that instead of reading from the input tape, it will
        read from the work tape. The new machine will run as follows: Run the new $M_g$ on $x$, and then run the new
        $M_f$ on what is written on the work tape. Correctness is trivial, we simply need to understand the space
        complexity. Let us call the combined machine $M$. We showed last week that the output of $M$ is indeed $f \left(g \left(x\right)\right)$, but the
        problem is that $M$ may use space that is not bounded by $O \left(\log \left(n\right)\right)$. This is because
        we showed last week that a machine with a limit of linear space, will have the runtime, and the output space
        bounded by a polynomial. This means that $\left|w\right|= \left|g \left(x\right)\right|$ may be polynomial, and
        so the machine $M$ may use polynomial space in $n$, instead of logarithmic in $n$. \\

        We will show a solution that allows the function assembly, such that in total the used space will be $O
        \left(\log \left(n\right)\right)$. .The concept is that we will compute the letters of $x$ dynamically, every
        time from the beginning. That is to say, we will run $M_f$, and $M_g$ simultaneously. We will hold 2 counters, one
        that changes the print head of $M_g$, and one that changes the print head of $M_f$ . The machine $M$ will
        compute $f \left(g \left(x\right)\right)$ as follows: It will run $M_f$. With every run step, in order to know
        what is the current letter of $g \left(x\right)$, $M$ will run $M_g$ the required number of prints, which is to
        say, for every step left of the read head of $f$, we will simulate this by reducing the relevant counter by 1,
        and every step right will increase it by one. Every instruction print in $M_g$ will be simulated by increasing
        the counter of the output of $M_g$. Every step of $M_f$, we will run $M_g \left(x\right)$ from the beginning
        (zeroing the print counter). We will run $M_g$ until the print counter is equal to the counter for the input
        tape of $M_f$. Then, $M$ will know the input of the letter from what $M)g$ prints, and $M_f$ reads. \\
        The
        correctness is trivial. \\
        The runtime is very large. We want to convince ourselves that the space complexity is $O \left(\log
        \left(n\right)\right)$. What does $M$ need to hold on the work tape? \begin{enumerate}
            \item Space of the work tape of $M_f$
            \item Space of the work tape of $M_g$
            \item Counter for the head of $M_g$
            \item Counter for the head of $M_f$
        \end{enumerate}
        \begin{enumerate}
            \item $M_f$ is contained in space $O \left(\log \left(\left|g \left(x\right)\right|\right)\right)$, ie log
                the length of output of $g \left(x\right)$. Reminder, a DTM that is contained in logarithmic space, has
                output contained by a polynomial, which is to say, length $\leq n^c$, where $c$ is some constant. That
                is to say, $M_f$ is contained in space $O \left(\log \left(n^c\right)\right)$. That is to say that $M_f$
                is contained in space $O \left(\log \left(n\right)\right)$.
            \item $M_g$ is contained in $O \left(\log \left(\left|x\right|\right)\right)$, in order to compute $g
                \left(x\right)$, which is to say $O \left(\log \left(n\right)\right)$
            \item We need counters that can count until $\left|g \left(x\right)\right|$, the length of the output of
                $M_g$, which is to say the length of the input of $M_f$. We have already shown that $\left|g
                \left(x\right)\right|$ is constrained by $n^c$, and therefore for a counter, we need $O \left(\log
                \left(n^c\right)\right)$ cells, which is $O \left(\log \left(n\right)\right)$.
            \item As above, and so below
        \end{enumerate}
        In total, we have $O \left(\log \left(n\right)\right)$ cells, which is to say that $M$ is a machine that
        computes $f \left(g \left(x\right)\right)$ in $O \left(\log \left(n\right)\right)$ space. Therefore, the
        assembly of two logarithmic space functions, is itself a logarithmic space computation.

        Note: We have shown that this is true for logarithmic space. It is also possible to show for other sub linear
        functions (but not all of them).
    \end{proof}
\end{theorem}
% section Sub linear space complexity (end)

\section{Classes L and NL}\label{sec:Classes L and NL} % (fold)
We will define a linear space reduction (we will stop reminding that this refers to space, and just say reduction for
this lecture). For the languages $A, B$, we will say that there is a linear reduction from $A$ to $B$, by writing \[
    A \leq_L B
\]
\textbf{if and only if} there exists a reduction $f$ from $A$ to $B$ such that $f$ is computable in linear space. \\
Note: If $A \leq_L B$ then $A \leq_p B$. (The machine that computes the reduction in logarithmic space will also compute
in linear time).  \\
The theorem of $L = NL$ is an open question, as is the theorem $L = P$. Additionally, so is the theorem $NL = P$. \\
We have seen that $L \subseteq P$, and we will show in the tutorial that $NL \subseteq P$.

\begin{theorem}[Closure of reduction assembly]
    \[
        A \leq_L B \land B \leq_L C \implies A \leq_L C
    \]
    \begin{proof}[Proof ]
        The proof is similar to the proof in the polynomial case. We use the closure of computable functions in
        logarithmic space. That is to say, if $f$ is computable in logarithmic space, and is the reduction from $A$ to
        $B$, and $g$ is computable in logarithmic space, and is the reduction from $B$ to $C$, then $g \circ f$ is
        computable in logarithmic space, and us the reduction from $A$ to $C$.
    \end{proof}
\end{theorem}

\begin{definition}[]
    We will say that the language $B$ is NL-hard, if for every $A \in NL$, it is true that $A \leq_L B$.
\end{definition}

\begin{theorem}[]
    If $B$ is a language that is NL-hard, and also $B \leq_L C$, then $C \in $NL-hard
    \begin{proof}[Proof ]
        Since $B \in$ NL-hard, then it is true that for every $A \in NL$ that $A \leq_L B$. Additionally, $B \leq_L C$
        (given). Closure of logarithmic space reductions is given, so the assembly gives $A \leq_L C$. That is to say,
        $C \in $NL-hard
    \end{proof}
\end{theorem}

\begin{definition}[]
    We will say that $C \in$ NL-Complete \textbf{if and only if} $C \in $NL-hard, and also $C \in NL$.
\end{definition}

In order to show that a language is complete in NL, will will do it similarly to Cook-Levin theorem for showing
completeness in NP.

Let us define \[
    \text{PATH}  = S-T-CONN = \left\{\left(G, s, t\right) : G \text{Is a directed graph } s, t \in V \left(G\right)
    \text{ and there is a directed path from $s$ to $t$ in } G\right\}
\]

\pagebreak
\begin{theorem}[]
    \[
        \text{PATH}  \in \text{NL-Complete}
    \]
    \begin{proof}[Proof ]
        We will show that \begin{enumerate}
            \item PATH $\in$ NL
            \item PATH $\in$ NL-hard
        \end{enumerate}

        \begin{enumerate}
            \item We will create an NTM, that is contained in space $O \left(\log \left(n\right)\right)$, for the
                language PATH. The machine $M$ will run as follows: Hold a variable for the current node, initialised to
                $s$. For every step that $M$ will take, in a non deterministic manner, it will go over the current nodes
                neighbours, and check if it has reached $t$. If so, then $M$ will stop, and return $q_{\text{acc}} $.
                Otherwise, it will continue to the next neighbour. \\
                \textbf{Correctness}: If $\left(G, s, t\right)\in $PATH, then there is a directed path from $s$ to $t$,
                in $G$, and therefore there is a computation where $M$ finds the path from $s$ to $t$, and so returns
                $q_{\text{acc}} $. \\
                If $\left(G, s, t\right) \notin$ PATH, then in every run of $M$ the path will never reach $t$, and
                therefore $M$ will not return $q_{\text{acc}} $. \\
                \textbf{Space}: $M$ uses the space on the work tape for the current node $O \left(\log
                \left(n\right)\right)$, and space for the next node. Another counter is used to find the edges in the
                representation of $G$ on the input tape in space $O \left(\log \left(n\right)\right)$. In total, we have
                used $O \left(\log \left(n\right)\right)$ space.

                Note: We can change $M$ such that it will always stop, for example, add a counter for the length of the
                path, which we increase for every step we take along the path, and if we reach the number of nodes in
                $G$, we stop and return $q_{\text{rej}} $.

                Note: The move from a node, to one of its neighbours, is through the input $G$. The way that this is
                done is dependent on the representation of $G$ (edges list, neighbours matrix, etc.). I any case, we may
                do this with one or two counters, which will take the additional space $O \left(\log
                \left(n\right)\right)$.
            \item We will now show that $\text{PATH} \in $NL-hard. That is to say, for every $A \in NL$, it is true that
                $A \leq_L $PATH. Or in other words, there is a function $f$, computable in logarithmic space, such that
                for every $w$, it is true that \[
                    w \in A \Leftrightarrow f \left(w\right) = \left(G, s, t\right) \in \text{PATH}
                \]
                The only additional given is that there exists a TM $M$, non deterministic, that is computable in
                logarithmic space for $A$.

                Reminder: The configurations graph $G_{M, w}$ (discussed last week) holds: There is a path from an
                initial configuration $c_0$, to the final configuration $c_{\text{acc}} $ in the graph $G_{M, w}$
                \textbf{if and only if} there is an accepting run of $M \left(w\right)$, \textbf{if and only if} $w \in
                A$. (W.l.o.g there is a single accepting configuration). \\
                Therefore, For the input $w$, the reduction will return $\left(G_{M, w}, c_0, c_{\text{acc}} \right) =
                \left(G, s, t\right)$. \\
                \textbf{Correctness}: Trivial. \\
                \textbf{Space complexity}: We need to show that we may compute this reduction in logarithmic space. We
                want to show that given $w$, to create $\left(G_{M, w}, c_0, c_{\text{acc}} \right)$ by a DTM, in
                logarithmic space. \\
                $c_0$ is the starting configuration, which is to say the work tape is empty, the heads are at the
                beginning, and the internal state is $q_0$. \\
                $c_{\text{acc}} $ is hte accepting configuration, meaning that the for both variables. working tape is empty, the head is at
                the start, and the internal state is $q_{\text{acc}} $. \\
                It is easy to create $c_0, c_{\text{acc}} $. It simply remains to clarify how we print $G_{M, w}$ in
                logarithmic space. We will choose the edges representation, which is to say, pairs of nodes. The
                reduction machine $M_f$ will create $G_{M, w}$ as follows: \\
                $M_f$ will hold a pair of variables on the working tape. Each one will representation a configuration.
                The required space to represent a configuration is \[
                    \left|Q\right| \cdot n \cdot O \left(\log \left(n\right)\right) \cdot \left|\Gamma\right|^{O
                    \left(\log \left(n\right)\right)}
                \]
                Which is the state, times the location of the input head, times the location of the working head, times
                the contents of the working tape. \\
                We may store one configuration in $O \left(\log \left(n\right)\right)$. For every pair of
                representations of the two variables, the machine will check if they are twinned to the pair of the
                following configuration. If so, it will copy / paste to the output tape. In every case, and if not, it
                will increment the variables, and the counters. That is to say, $M$ will pass over all the possible
                values, for every pair of variables.  To check if a pair of values is suitable to a pair of following
                configurations, which is to say if 2 configurations are almost equal, for example, changes that twin to
                the $\delta$ function of $M$, are able to be done in the space limitations. \\
                Note: In order to save a configuration, we need the space for the size of the work tape, $O \left(\log
                \left(n\right)\right)$, and the location of the head of the work tape $O \left(\log \left(\log
                \left(n\right)\right)\right)$, the head for the input tape $O \left(\log \left(n\right)\right)$, and
                internal state $O \left(\log \left(Q\right)\right)$. In total, $O \left(\log \left(n\right)\right)$.
        \end{enumerate}

    \end{proof}
\end{theorem}

We have / will show \[
    L \subseteq NL \subseteq P \subseteq NP
\]
We will show the middle containment in the tutorial. We are left with the following questions: \begin{itemize}
    \item $L = NL$
    \item $L = P$
    \item $NL = P$
    \item $P = NP$
\end{itemize}
These are still open questions, to which the world knows not the answers.
% section Classes L and NL (end)

\end{document}
