\documentclass[a4paper]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{blindtext}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{booktabs}
\usepackage{float}
\usepackage{bbm}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows}
% \usepackage{stix}
\graphicspath{ {./images/} }
\usepackage[bottom=0.5cm, right=1.5cm, left=1.5cm, top=1.5cm]{geometry}

\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{exercise}{Exercise}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}[section]

\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}
\newcommand{\p}{\ensuremath{\mathbb{P}}}
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\1}{\ensuremath{\mathbbm{1}}}
\newcommand{\B}{\ensuremath{\mathbbm{B}}}

\title{Tutorial 12}
\author{Gidon Rosalki}
\date{2025-06-18}


\begin{document}
\maketitle
\section{The Time Hierarchy Theorem}\label{sec:The Time Hierarchy Theorem} % (fold)
\begin{definition}[Time-constructible]
    A function $t: \N \to \N$ where $t \left(n\right) = \Omega \left(n \log \left(n\right)\right)$ is called
    \textbf{time-constructible} of the function that maps $1^n$ ($n$ in unary form) to the binary representation of $t
    \left(n\right)$ is computable in time $O \left(t \left(n\right)\right)$
\end{definition}

Note: All non linear polynomials with integer coefficients and non-negative leading coefficient are time-constructible,
so are exponential functions such as $2^n$. \\

\begin{example}[]
    Is the function $n^2$ time-constructible?
    \begin{proof}[Solution]
        Given $n$ 1s written on the tape, we need to write $n^2$ 1s. We will use the following algorithm: \begin{enumerate}
            \item Translate the input $1 \dots 1$ to the binary representation of $n$, twice
            \item Multiply the two numbers together by "vertical multiplication"
        \end{enumerate}
        Time complexity: \begin{enumerate}
            \item $n \log \left(n\right)$
            \item $\log \left(n\right)^2$
        \end{enumerate}
        In total, this is $O \left(n^2\right)$, so this function is in fact time-constructible.
    \end{proof}
\end{example}

Reminder: $f \left(n\right) = o \left(g \left(n\right)\right) \implies \displaystyle\lim_{n \to \infty}
\displaystyle\frac{f \left(n\right)}{g \left(n\right)} = 0$
We also have the following time hierarchy theorem, which we state here without a proof:
\begin{theorem}[Time Hierarchy]
    Let $t: \N \to \N$ be a time constructible function. Then there exists a language $L$, that is decidable in $O
    \left(t \left(n\right)\right)$ time, but not decidable in $o \left(\displaystyle\frac{t \left(n\right)}{\log \left(t
    \left(n\right)\right)}\right)$ time
\end{theorem}

\begin{corollary}[]
    For any two real numbers $1 \leq \varepsilon_1 < \varepsilon_2$, we have TIME $\left(n^{\varepsilon_1}\right)
    \subset$ TIME $\left(n^{\varepsilon_2}\right)$
\end{corollary}

\begin{corollary}[]
    $P \subset $ EXPTIME
    \begin{proof}[Proof ]
        For every $k$, it holds that $n^k = O \left(2^n\right)$, so TIME $\left(n^k\right) \subset $ TIME
        $\left(2^n\right)$, and therefore $P \subset $ TIME $\left(2^n\right)$. By the time hierarchy theorem, we know
        that \[
            \text{TIME}  \left(2^n\right) \subsetneq \text{TIME} \left(4^n\right) \subset \text{EXPTIME}
        \]
    \end{proof}
\end{corollary}
% section The Time Hierarchy Theorem (end)

\section{The Space Hierarchy Theorem}\label{sec:The Space Hierarchy Theorem} % (fold)
While with respect to time complexity simulation comes with a logarithmic overhead, the same is not true for space and
so the analogous space hierarchy theorem is cleaner, and provides a tighter bound.

\begin{definition}[Space-constructible]
    A function $s: \N \to \N$ where $s \left(n\right) = \Omega \left(n\right)$ is called \textbf{space-constructible} if
    the function that maps $1^n$ ($n$ in unary form) to the binary representation of $s \left(n\right)$ is computable in
    space $O \left(s \left(n\right)\right)$
\end{definition}
\begin{theorem}[Space Hierarchy]
    Let $s: \N \to \N$ be a space-constructible function. Then there exists a language $L$, that is decidable in $O
    \left(s \left(n\right)\right)$ space, but not decidable in $o \left(s \left(n\right)\right)$ space
\end{theorem}

\begin{corollary}
    SPACE $\left(n^k\right) \subsetneq $ SPACE $\left(n^{k + 1}\right)$ for all $k \geq 1$, and PSPACE $\subsetneq$
    EXPSPACE
\end{corollary}
% section The Space Hierarchy Theorem (end)

\section{Padding}\label{sec:Padding} % (fold)
\begin{theorem}[The following are equivalent]
    \begin{enumerate}
        \item PSPACE = PTIME
        \item $\exists k > 1 : \text{SPACE} \left(n^k\right) \subseteq \text{PTIME} $
        \item SPACE $\left(n\right) \subseteq $ PTIME
    \end{enumerate}

    \begin{proof}[Proof ]
        It is clear that if PSPACE = PTIME, then $\text{SPACE} \left(n^k\right) \subseteq \text{PSPACE} = \text{PTIME} $
        for all $k \geq 1$. Also, if SPACE $\left(n^k\right) \subseteq $ PTIME, for some $k \geq 1$, then \[
            \text{SPACE} \left(n\right) \subseteq \text{SPACE} \left(n^k\right) \subseteq \text{PTIME}
        \]
        It is left to prove that if SPACE $\left(n\right) \subseteq $ PTIME, then PSPACE = PTIME, and for this we use
        the padding argument. \\
        Assume that SPACE $\left(n\right) \subseteq $ PTIME, and consider a language $L \in PSPACE$. Let $f
        \left(n\right) = n^k$ be such that $L \in \text{DSPACE} \left(f \left(n\right)\right)$. Consider the language \[
            L' = \left\{w\#1^m : w \in L \land m = \left|w\right|^k\right\}
        \]
        It is not hard to see that $L' \in \text{SPACE } \left(n\right) \subseteq \text{PTIME} $. Indeed, $f
        \left(n\right)= n^k $ is space constructible, and so we can first compute $\left|w\right|^k$, as long that the
        space for the output does not exceed $m$ cells (and if it does, we will reject), and then compare
        $\left|w\right|^k$ with $m$. If they are not equal, then we reject, and otherwise, if $m = \left|w\right|^k$,
        then we decide whether $w \in L$ using \[
            O \left(f \left(\left|w\right|\right)\right) = O \left(\left|w\right|^k\right) = O
            \left(\left|w\#1^m\right|\right) = O \left(n\right)
        \]
        space. Thus, $L' \in \text{PTIME} $. It then follows that $L \in \text{PTIME} $: On input $w$, we first compute
        $m = \left|w\right|^k$, in polynomial time in $\left|w\right|$, and then determine whether $w\#1^m \in L'$. The
        latter can be done in polynomial time in $\left|w\#1^m\right| = O \left(\left|w\right|^k\right)$, which is
        polynomial in $\left|w\right|$. Hence, $L \in \text{PTIME} $, and we are done.
    \end{proof}
\end{theorem}

\begin{corollary}
    For all $k \geq 1$, it holds that $\text{SPACE} \left(n^k\right) \ne \text{PTIME} $
    \begin{proof}[Proof ]
        By the previous lemma, $\text{SPACE} \left(n^k\right) = \text{PTIME} $ implies PSPACE = PTIME, which clearly
        implies that \[
            \text{SPACE} \left(n^{k  +1}\right) \subseteq \text{PSPACE} = \text{PTIME} = \text{SPACE} \left(n^k\right)
        \]
        The latter is a contradiction to the separation \[
            \text{SPACE} \left(n^k\right) \subsetneq \text{SPACE} \left(n^{k + 1}\right)
        \]
        given by the space hierarchy theorem
    \end{proof}
\end{corollary}
% section Padding (end)

\section{Logarithmic space languages}\label{sec:Logarithmic space languages} % (fold)
\begin{definition}[Logarithmic Space TM]
    A logarithmic space TM is a TM with 3 tapes: \begin{enumerate}
        \item Input tape: Read only tape, containing letters only from $\Sigma$
        \item Working tape: May write any $\sigma \in \Gamma$, and utilise $O \left(\log \left(n\right)\right)$ space
        \item Output tape: This tape is write only, and may have any letter $\sigma \in \Sigma$ written to it, and it
            may only move right, or halt.
    \end{enumerate}
\end{definition}
We may now define the following languages: \begin{gather*}
    L = \text{SPACE}  \left(O \left(\log \left(n\right)\right)\right) \\
    NL = \text{NSPACE}  \left(O \left(\log \left(n\right)\right)\right)
\end{gather*}
Remember your laws of logarithms, $\log \left(n^k\right) = k \log \left(n\right)$. Thus, for every polynomial $p$, it is
true that $\log \left(p \left(n\right)\right) = O \left(\log \left(n\right)\right)$. This is why we do not take powers
of $n$ in the log. However, note that $\log \left(n\right) = o \left(\log^2 \left(n\right)\right)$. So a machine that
works in $\log^2 \left(n\right)$ does not belong to $L$. \\
We have seen that \[
    \text{SPACE}  \left(f \left(n\right)\right) \subseteq \text{TIME} \left(2^{O \left(f \left(n\right)\right)}\right)
\]
Thus, if $f = O \left(\log \left(n\right)\right)$, then \[
    n \cdot 2^{O \left(f \left(n\right)\right)} = n \cdot 2^{O \left(\log \left(n\right)\right)} = n \cdot poly \left(n\right)
\]
We can conclude that $L \subseteq P$, and next week we will show that $NL \subseteq P$
% section Logarithmic space languages (end)

\section{NL-Completeness}\label{sec:NL-Completeness} % (fold)
Similarly to the question of whether P = NP, we do not know whether L = NL. As we did in NP, we want to say on some
problems in NL that they are the "hardest". Thus, we want to define the notion of NL-hardness. As we did in NP, we want
to say that a problem is NL-hard if every other problem in NL is reducible to it. What kind of reductions should we use?
This is a very important point, yet difficult to understand.

\subsection{Log space reductions}\label{sub:Log space reductions} % (fold)
\begin{definition}[Log space reducible]
    We say that a language that $A$ is log-space reducible to $B$ if there exists a logarithmic space computable
    function $f : \Sigma^* \to \Sigma^*$ such that $\forall w,\ w \in A \Leftrightarrow f \left(w\right) \in B$. In this
    case we will say $A \leq_L B$.
\end{definition}

\begin{definition}[NL-hard]
    We say that a language $L$ is NL-hard if $\forall L' \in NL,\ L' \leq_L L$. If in addition, $L \in NL$, then we say
    that $L$ is NL-complete
\end{definition}

\begin{theorem}[Reduction theorem for LOGSPACE]
    If $B \in L$, and $A \leq_L B$, then $A \in L$. If $B \in NL$, and $A \leq_L B$, then $A \in NL$. Similarly, if $A
    \notin LNL$, and $A \leq_L B$, then $B \notin LNL$
\end{theorem}

\begin{corollary}
    If $A \in$ NL-complete, and $A \in L$, then L = NL
\end{corollary}

\begin{theorem}[]
    If $A \leq_L B$, and $B \leq_L C$, then $A \leq_L C$

    \begin{proof}[Proof ]
        Left as an exercise, it is similar to the proof of log space reductions
    \end{proof}
\end{theorem}
% subsection Log space reductions (end)

\subsection{Relations between classes}\label{sub:Relations between classes} % (fold)
So far we have seen the following relations between classes. \[
    L \subseteq NL \subseteq P \subseteq NP \subseteq \text{PSPACE} = \text{NPSPACE}  \subseteq EXP
\]
From the hierarchy theorems, we thus get that \[
    L \subsetneq \text{PSPACE},\ P \subsetneq \text{EXP}
\]
And a non deterministic version of the hierarchy theorem shows that \[
    NL \subsetneq \text{NPSPACE}  = \text{PSPACE}
\]
% subsection Relations between classes (end)
% section NL-Completeness (end)

\section{STRONGLY-CONNECTED is NL-Complete}\label{sec:STRONGLY-CONNECTED is NL-Complete} % (fold)
A directed graph is strongly connected if every two nodes are connected by a directed path in each direction. We define
the language \[
    \text{STRONGLY-CONNECTED}  = \left\{\left\langle G \right\rangle : G \text{ is a strongly connected graph} \right\}
\]
and we claim that it is NL-complete. First, we will show that it is in NL, to do this we will use the Immerman theorem,
which we will prove next week.

\begin{theorem}[Immerman]
    NL = coNL
\end{theorem}

We proved in the lecture that \[
    \text{PATH}  = \left\{\left\langle G, s, t \right\rangle : G \text{ is a directed graph, and there exists a path
    from $s$ to $t$} \right\}
\]
is NL-complete. As such, $\overline{\text{PATH} }$ is NL-complete. So $\overline{\text{PATH} } \in NL$. Thus, there
exists an NTM, that given $\left\langle G, s, t \right\rangle$, accepts \textbf{if and only if} there is no path from
$s$ to $t$, in $G$. Consider the following machine which recognises $\overline{\text{STRONGLY-CONNECTED} }$. \\
Given an input $\left\langle G \right\rangle$: \begin{enumerate}
    \item Non-deterministically select two nodes $a$, and $b$
    \item Simulate the NTM that decides $\overline{\text{PATH} }$ on input $\left\langle G, a, b \right\rangle$. If it
        accepts, then the graph is not strongly connected, so accept. Otherwise, reject
\end{enumerate}
Since storing node numbers $a$ and $b$ only takes logarithmic space, and $\overline{\text{PATH} }$ is recognised in
logarithmic space, we get that $\overline{\text{STRONGLY-CONNECTED} } \in NL$. Again, from Immerman, we get that
$\text{STRONGLY-CONNECTED}  \in NL$. \\
Next, we show that every other language in NL is log-space reducible to STRONGLY-CONNECTED. We do this by reducing PATH
(which is NL-complete) to STRONGLY-CONNECTED. Consider the following reduction: \\
On input $\left\langle G, s, t \right\rangle$, where $G = \left(V, E\right)$, the reduction outputs $\left\langle G'
\right\rangle$, where $G'$ is obtained from $G$ by adding, for every $v \in V$, the edges $\left(v, s\right)$ and
$\left(t, v\right)$. \\
We will claim that the reduction is correct, and that it can be done in log-space. \\

\textbf{Correctness}: We need to show that if there is a path from $s$ to $t$ in $G$, then $G'$ is strongly connected.
For $x, y \in V$ a path from $x$ to $y$ starts with the edge $\left(x, s\right)$, and then takes the path from $s$ to
$t$, and then finally takes the edge $\left(t, y\right)$. \\
Conversely, we need to show that if there is no path from $s$ to $t$, then $G'$ is not strongly connected. Indeed, we
claim that $t$ is not reachable from $s$ in $G'$, since a simple path from $s$ to $t$ in $G'$ cannot use any of the new
edges, and thus it is also a path in $G$. Since the only additional edges in the constructed graph go into $s$, and out
of $t$, there can be no new ways of reaching $t$ from $s$. \\

\textbf{Space}: We need to verify that the reduction can be performed by a log-space TM. A log-space TM can start by
copying the input $G$, and for every vertex it writes, it also writes the two new edges. This is done vertex by vertex,
and thus we always keep at most the encoding of a vertex, or an edge, on the tape. Both of these are logarithmic in the
input. So the reduction TM will do the following: \begin{enumerate}
    \item Copy all of $G$ onto the output tape
    \item For  each node $i$ in $G$: \begin{enumerate}
        \item Add an edge from $i$ to $s$
        \item Add an edge from $t$ to $i$
    \end{enumerate}
\end{enumerate}
So, the space used is indeed logarithmic, although the output has size $O \left(n\right)$, essentially the only space
necessary to perform the reduction is that used to keep track of the node number $i$ in the above for loop.
% section STRONGLY-CONNECTED is NL-Complete (end)

\end{document}
