\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{blindtext}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{booktabs}
\usepackage{float}
\usepackage{bbm}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows}
% \usepackage{stix}
\graphicspath{ {./images/} }
\usepackage[bottom=0.5cm, right=1.5cm, left=1.5cm, top=1.5cm]{geometry}

\newtheorem{theorem}{Theorem}
\newtheorem{exercise}{Exercise}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}[section]

\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}
\newcommand{\p}{\ensuremath{\mathbb{P}}}
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\1}{\ensuremath{\mathbbm{1}}}
\newcommand{\B}{\ensuremath{\mathbbm{B}}}

\title{Tutorial 7}
\author{Gidon Rosalki}
\date{2025-05-14}


\begin{document}
\maketitle
\section{Non deterministic Turing machines}\label{sec:Non deterministic Turing machines} % (fold)
\begin{definition}[Nondeterministic Turing Machine - NTM]
    A non-deterministic Turing Machine is a 7-tuple \[
        M = \left\langle \Sigma, \Gamma, \text{\textvisiblespace}, Q, Q_0, F, \delta \right\rangle
    \]
    Where: \begin{itemize}
        \item $\Sigma$ is a finite alphabet
        \item $\Gamma$ is a finite tape alphabet, such that $\Sigma \subseteq \Gamma$
        \item \textvisiblespace $\in \Gamma \setminus \Sigma$ is the blank symbol
        \item $Q$ is a finite, non empty, set of states
        \item $Q_0 \subseteq Q$ is the set of initial states
        \item $F \subseteq Q$ is the set of final states
        \item $\delta : \left(Q \setminus F\right) \times \Gamma \to 2^{Q \times \Gamma \times \left\{L, R\right\}}$ is
            the transition function
    \end{itemize}
    The transition function δ maps a state and a tape symbol to a set of possible actions, where each action consists of
    a next state, a symbol to write, and a direction to move the tape head (left or right).
\end{definition}

\begin{definition}[Acceptance]
    Let $N$ be an NTM and let $w \in \Sigma^*$. We say that $N$ accepts $w$ if there exists at least one accepting run
    of $N$ on $w$ - that is, a run that ends in $q_{\text{acc} }$. In this case, we write $w \in L \left(N\right)$.
\end{definition}

\begin{example}[CLIQUE]
    Let $G = \left(V, E\right)$ be a graph. A \textbf{clique} is a set $S \subseteq V$uch that there is an edge between
    every pair of vertices in $S$. Define the following decision problem: \[
        CLIQUE = \left\{\left\langle G, k \right\rangle : \text{There is a clique of size } k \text{ in } G\right\}
    \]

    \begin{proof}[Solution]
        We will describe an NTM that decides $CLIQUE$. The machine non-deterministically writes on the tape the names of
        $k$ distinct vertices from $V$, separated by $\#$. This guessing is done by replacing a blank section of the
        tape with symbols from the encoding of vertex names, one at a time, using non-deterministic transitions. Then,
        the machine checks deterministically whether all $\binom{k}{2} $ pairs among the chosen vertices are connected
        by edges in G. For each pair, it searches through the encoding of E to see if the corresponding edge exists. \\
        If all pairs are connected, then the machine accepts, otherwise it rejects. \\
        This machine runs in non-deterministic, polynomial time. Guessing the $k$ vertex names takes polynomial time,
        since $k \leq \left|V\right|$. Checking all $\binom{k}{2} $ pairs for the existence of pairs takes polynomial
        time, by scanning the input encoding of $G$. Therefore, $CLIQUE$ can be solved by an NTM in polynomial time.
    \end{proof}
\end{example}

We shall however note, that a trivial deterministic algorithm would try all subsets of $k$ vertices and check if any of
them forms a clique. However, the number of such subsets is $\binom{\left|V\right|}{k}$, which is exponential in the
input size when $k$ is not a constant because: \[
    \binom{\left|V\right|}{\frac{\left|V\right|}{2}} \approx 2^{\left|V\right|}
\]
Since each check takes polynomial time but must be repeated exponentially many times, this yields an exponential-
time algorithm overall. Thus, the trivial approach does not give a polynomial-time solution.
% section Non deterministic Turing machines (end)

\section{P and NP}\label{sec:P and NP} % (fold)
\begin{definition}[Runtime of a TM]
    The \textbf{runtime} of a TM $M$ on an input $w \in \Sigma^*$ is the number of steps that $M$ performs before it
    halts.
\end{definition}

\begin{definition}[Runtime of an NTM]
    The \textbf{runtime} of an NTM $N$ on an input $w \in \Sigma^*$, is the number of steps that $N$ performs, before
    halting, in the longest run on $w$.  \[
        runtime_N \left(w\right) = \displaystyle\max_{} \left\{\left|r\right| : r \text{is a run of } N \text{ on } w\right\}
    \]
\end{definition}

\begin{definition}[Runtime complexity]
    For $f: \N \to \N$, we say that a TM/NTM $T$ runs in time $O \left(f \left(n\right)\right)$ if \[
        \forall w \in \Sigma^*\ runtime_T \left(w\right) \in O \left(f \left|w\right|\right)
    \]
\end{definition}

\begin{definition}[]
    For $f: \N \to \N$, \begin{align*}
        TIME \left(f \left(n\right)\right) &= \left\{L \subseteq \Sigma^* : L \text{ can be decided by a TM that runs in
        } O \left(f \left(n\right)\right)\right\} \\
        NTIME \left(f \left(n\right)\right) &= \left\{L \subseteq \Sigma^* : L \text{ can be recognised by an NTM that
        runs in } O \left(f \left(n\right)\right)\right\}
    \end{align*}
\end{definition}

\begin{definition}[P/NP]
    The class of languages that can be decided in polynomial time, by the relevant TM/NTM: \begin{gather*}
        P = \displaystyle\bigcup_{k = 1}^{\infty}TIME \left(n^k\right) \\
        NP = \displaystyle\bigcup_{k = 1}^{\infty}NTIME \left(n^k\right)
    \end{gather*}
\end{definition}
There is of course, the million dollar question, does $P = NP$? The proof is left as an exercise for the reader (note,
finding a proof, or disproof, will result in getting 100\% in the course, and a million dollars, but that is far less
interesting).

\begin{definition}[EXP]
    The class of languages that can be decided in exponential time by a TM \[
        EXP = \displaystyle\bigcup_{k = 1}^{\infty} \left(2^{n^k}\right)
    \]
\end{definition}

If a language can be decided by a TM that runs in polynomial or exponential time, then in particular it is decidable.
Therefore it follows easily from the definition that $P, EXP \subseteq R$. Moreover, in the next homework we will show
that $NP \subseteq EXP$, (by describing a TM that simulates all possible runs of the NTM), thus also $NP \subseteq R$

\subsection{Closure properties of P/NP}\label{sub:Closure properties of P/NP} % (fold)
\begin{theorem}[$P$ is closed under union]
    $P$ is closed under union:
    \begin{proof}[Proof ]
        Let $L_1, L_2 \in P$. Then there are 2 TMs, $M_1, M_2$, that decide $L_1, L_2$ respectively, and run in
        polynomial time. Let us define a new TM $M$, which operates as follows on the input $x \in \Sigma^*$: \begin{itemize}
            \item Run $M_1$ on $x$, and accept if it accepts
            \item Run $M_2$ on $x$, and accept if it accepts
            \item If neither accepted, then reject
        \end{itemize}
        The proof that $L \left(M\right) = L_1 \cup L_2$ is identical to the case of union in $R$ (shown in a previous
        recitation). The runtime of $M_1$ and $M_2$ is polynomial, so the sum is also polynomial.  \\
        We will also note that during the run of $M_1$ on $x$, the machine $M$ must keep a copy of $x$ for later (to
        clean up the tape for $M_2$). This involves overhead: after each step that $M_1$ performs, $M$ may have to move
        the copy. The cost is polynomial in $\left|x\right|$, per step. So the overall runtime of M is also polynomial.
    \end{proof}
\end{theorem}

\begin{theorem}[NP is closed under union]
    NP is closed under union
    \begin{proof}[Proof ]
        Let $L_1, L_2 \in NP$. Then there are 2 TMs, $N_1, N_2$, that recognise $L_1, L_2$ respectively, and run in
        polynomial time. Let us define a new TM $N$, which recognises $L_1 \cup L_2$ as follows: \begin{itemize}
            \item On the input $w$, $N$ will choose non-deterministically whether to simulate $N_1$ or $N_2$
            \item It then simulates the chosen machine on input $w$
            \item If the simulation accepts, then $N$ accepts, otherwise it rejects
        \end{itemize}

        Since both $N_1$ and $N_2$ run in polynomial time, and the non-deterministic choice adds only a constant
        overhead, the machine N also runs in non-deterministic polynomial time (The maximum of the time of $N_1$ and
        $N_2$). \\
        Moreover, $N$ accepts $w$ \textbf{if and only if}  $w \in L_1 \lor w \in L_2$, so $L \left(N\right) = L_1 \cup
        L_2 \in NP$
    \end{proof}
\end{theorem}

\begin{theorem}[P is closed under complement]
    \begin{proof}[Proof ]
        Let $L \in P$, and let $M$ be a deterministic TM that decides $L$ in polynomial time. We shall construct a
        machine $M'$ by swapping the accepting and rejecting states of $M$. Since this transformation only modifies the
        final states and leaves the computation unchanged, $M'$ still runs in polynomial time and decides the complement
        language $L$. Hence, $L \in P$.
    \end{proof}
\end{theorem}

This proof does not extend to NP. Consider CLIQUE, and now consider the complement language \[
    \overline{CLIQUE} = \left\{\left\langle G, k \right\rangle : \text{Every size } k \text{ subsets of vertices is not
    a clique} \right\}
\]
To decide this language, we must confirm that \textbf{no} subset of size $k$ forms a clique. A non-deterministic machine
cannot verify this efficiently: it can guess a subset and reject if it is a clique, but if it guesses a non-clique, that
tells us nothing about the existence of a different subset that might be a clique. Thus, flipping accept and reject
states of the non-deterministic machine for $CLIQUE$ does not yield a valid machine for $\overline{CLIQUE}$, since
rejection does not mean "no" in the presence of non-determinism. In fact, it is currently unknown whether NP is closed
under complement; that is, whether $NP = coNP$ remains an open question.
% subsection Closure properties of P/NP (end)
% section P and NP (end)

\section{Graph problems}\label{sec:Graph problems} % (fold)
We shall encode graphs as follows: For a graph $G = \left(V, E\right)$, we denote by $\left\langle G \right\rangle$ its
encoding as a word over $\left\{0, 1, \#\right\}$. It contains a binary name for each vertex in $V$ , and a pair of such
names for each edge in $E$. Notice that this encoding is the same size as the graph $\left|V\right| + \left|E\right|$
(up to a polynomial).

\begin{example}[]
    Let $L = \left\{\left\langle G \right\rangle : G \text{ is a connected graph} \right\}$. We will show that $L \in P$
    (remember, we did this in DAST using DFS).
    \begin{proof}[Solution]
        \begin{itemize}
            \item Choose an arbitrary vertex $v_0 \in V$
            \item Perform DFS starting at $v_0$
            \item If all vertices are visited, then the graph is connected, otherwise it is not
        \end{itemize}
        The time complexity taught in DAST was $O \left(\left|V\right| + \left|E\right|\right)$. This does not hold for
        TMs, since every iteration (checking whether or not a node has been visited, and finding its neighbours),
        requires scanning the tape several times. The TM is not as efficient, but, importantly, the difference in
        efficiency is polynomial. This is because the main difference between TMs and the model we are used to (random
        access memory) is that the TM cannot "jump" to a memory cell, so we pay a (polynomial) price in time per step.
    \end{proof}
\end{example}

There are many important decision problems in graph theory, for example: \begin{itemize}
    \item An \textbf{independent set} in a graph $G = \left(V, E\right)$ is a subset $S \subseteq V$ such that no two
        vertices in $S$ are connected by an edge
    \item A \textbf{Hamiltonian path} is a path that visits every vertex in a graph \textit{exactly once}.
\end{itemize}
From these, we can define the following decision problems: \begin{gather*}
    IS = \left\{\left\langle G, k \right\rangle : \text{There is an independent set of size } k \text{ in } G\right\} \\
    HAM-PATH = \left\{\left\langle G \right\rangle : \text{There is a Hamiltonian path in } G\right\}
\end{gather*}
A natural question arises: which of these problems are in P , and which are in NP? How can we formally compare the
complexity of such problems? \\
To answer this, we introduce the notion of polynomial-time reductions, which help us classify problems, and understand
the relationships between them within the complexity hierarchy.
% section Graph problems (end)

\section{Polynomial reductions}\label{sec:Polynomial reductions} % (fold)
\begin{definition}[Computable in polynomial time]
    A function $f: \Sigma^* \to \Sigma^*$ is computable in polynomial time if there exists a TM $M_f$ that runs in
    polynomial time, and for every input $x \in \Sigma^*$, when $M_f$ halts, the word $f \left(x\right)$ is written to
    the tape
\end{definition}

\begin{definition}[Polynomial reduction]
    A polynomial reduction from a language $A \subseteq \Sigma^*$ to a language $A \subseteq \Sigma^*$ is a function $f
    : \Sigma^* \to \Sigma^*$ from $A$ to $B$ that is computable in polynomial time. If such a reduction exists, we
    denote $A \leq_p B$.
\end{definition}

A big reason we introduced reductions was for the reduction theorem, so let us extend it to this set of definitions:

\begin{theorem}[Reduction theorem for P]
    If $A \leq_p B$, and $B \in P$, then $A \in P$
    \begin{proof}[Proof ]
        Suppose $A \leq_p B$. There is a TM $M_f$ that computes the reduction in polynomial time. Since $B \in P$, there
        is some TM $M_B$ that decides $B$ in polynomial time. We define $M_A$ which runs as follows on input $x \in
        \Sigma^*$: \begin{itemize}
            \item Run $M_f$ on $x$ to obtain $f \left(x\right)$
            \item Run $M_B$ on $f \left(x\right)$, and accept or reject accordingly
        \end{itemize}
        Correctness is the same as the reduction theorem for mapping reductions. It remains to prove that this algorithm
        indeed runs in polynomial time. \\
        The output of $M_f \left(x\right)$ is of length $O \left(\left|x\right|^k\right)$ for some constant $k$, and the
        runtime of $M_B$ on an input of size $O \left(\left|x\right|^k\right)$ is $O
        \left(\left(\left|x\right|^k\right)^m\right)$ for some constant $m$. Therefore the running time is $O
        \left(\left|x\right|^{mk}\right)$, which is polynomial in $\left|x\right|$.
    \end{proof}
\end{theorem}

\begin{theorem}[]
    Consider the languages \begin{align*}
        CLIQUE &= \left\{\left\langle G, k \right\rangle : \text{There is a clique of size } k \text{ in } G\right\} \\
        IS &= \left\{\left\langle G, k \right\rangle : \text{There is an independent set of size } k \text{ in } G\right\} \\
    \end{align*}
    Prove that $CLIQUE \leq_p IS$
    \begin{proof}[Proof ]
        The idea is that cliques, and independent sets are complementary: we can convert between them by complementing
        the graph.
        \begin{itemize}
            \item Construction: Let $f \left(\left\langle G, k \right\rangle\right) = \left\langle \overline{G}, k
                \right\rangle$, where $\overline{G}$ is obtained from $G$ by removing the edges, and then adding edges
                between vertices that were not connected before, or formally: $\overline{G} = \left(V,
                \overline{E}\right)$ where \[
                    \overline{E} = \left\{\left\{u, v\right\} :  u \ne v \in G \land \left\{u, v\right\} \notin E\right\}
                \]
            \item Correctness: \begin{itemize}
                    \item Suppose $\left\langle G, k \right\rangle \in CLIQUE$. Then there is some set of vertices $S$ of
                        size $k$ such that there is an edge between every two vertices in $S$. By definition of $G$, there
                        are no edges at all between the vertices in $S$ in $G$, so it is an independent set of size $k$ and
                        we have $\left\langle G, k \right\rangle \in IS$
                    \item Suppose $\left\langle \overline{G}, k \right\rangle \in IS$. Then there exists a set $S \subseteq
                        V$ of size $k$ such that there are no edges between any two vertices in $S$. Thus $S$ is a clique in
                        $G$, so $\left\langle G, k \right\rangle \in CLIQUE$
                \end{itemize}
            \item Complexity: A TM can iterate over every $\left\{u, v\right\} \in V \times V$ , check if it’s in $E$
                and add or not add it to $G$ accordingly. There are $O \left(\left|V\right|^2\right)$ pairs to check,
                and each takes polynomial time (we iterate over the edges of $G$ and perform string comparisons). Thus
                the reduction can be computed in polynomial time
        \end{itemize}
    \end{proof}
\end{theorem}

We don’t know whether $CLIQUE$ and $IS$ are in $P$ or not. However now we know that if $IS \in P$ then $CLIQUE \in P$.
There is also a polynomial reduction in the other direction (see exercise 7). So they "go together": either both are in
$P$ or both are not in $P$. We will see that this situation applies to many other problems.

% section Polynomial reductions (end)


\end{document}
